Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-05-01 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
New results section and removed bullshit from the discussion

We want to know what the effect is of inducing non-conv of the bias in
all Qs and the confidence validity of the betas. So we plot the bias for
each Q against the number of iterations in the condition and the
missingness proportion in the condition. We look at averages across
simulation repetitions. Then, we want to know if conditions in which the
Qs are affected by non-convergence (on average) are also identified by
the conv diag. We do this by looking at the bias in Q against the

\textbf{Add to this section: 1) emphasize that the plots are averages
across repetitions, not within MICE; 2) sample effects due to single
complete dataset.}

For reasons of brevity, we only discuss the convergence diagnostics for
the \(Q\) with the worst performance in terms of bias. For the
descriptve statistics, the magnitude of the bias was the largest in
\(Y\) (i.e, \(j = Y\) in \(\mu_j\) and \(\sigma_j\)). For the
\(Q=\beta_j\), \(j=2\) showed the worst perfomance (i.e, the effect of
\(X_1\) on \(Y\).

@@

Evaluation with the performance measures shows that \(\widehat{R}\) and
autocorrelation are conservative: they indicate signs of non-convergence
in conditions where \(\bar{Q}\)s are unbiased and or confidence valid
estimates of \(Q\).

\hypertarget{old-results-section}{%
\subsection{Old Results section}\label{old-results-section}}

\textbf{(Add more info about figure legends and axes.)}

@@

@@

@@

@@

\hypertarget{univariate-estimates-and-convergence-diagnostics}{%
\subsection{Univariate estimates and convergence
diagnostics}\label{univariate-estimates-and-convergence-diagnostics}}

The bias in the estimates of the variable means shows little to no
difference between simulation conditions. It doesn't seem to matter how
many iterations you use in the mice algorithm, the estimates are
unbiased. Similarly, the bias in the estimated variances is more or less
stable across simulation conditions. These univariate quantities appear
to be unaffected by the number of iterations.

When applied to the imputation chain means, \(\widehat{R}\) indicates
that the mice algorithm does not reach a converged state
(\(\widehat{R}\) = 1) in any of the simulation conditions. Neither is
the most recent recommended threshold reached (\(\widehat{R}\)
\textless{} 1.01). \textbf{This is only for 75\% missingness:} The
conventional \(\widehat{R}\) threshold of 1.2 is reached in simulation
conditions \(T = 3\) and \(T > 6\). (With the default number of
iterations (maxit = 5), this dip in \(\widehat{R}\) values would be
spotted, so it is no problem.) The point at which an extra iteration
does not seem to improve the \(\widehat{R}\) value is around \(T=30\).

Autocorrelations indicate no sign of trending \textbf{(NOT correct
anymore, fluke of the stats::acf() function!!)} within imputation
chains. In most simulation conditions \(AC\) is smaller than or about
equal to zero. Across simulation conditions, however, the
autocorrelation curve does not trend towards 0. Autocorrelation values
plateau off at a value of around.1. This is a small positive
autocorrelation, which would indicate some trending within chains.

The \(\widehat{R}\) and \(AC\) values for the imputation chain variances
show equal trends to the chain means and are therefore not discussed
separately. Taken together, univariate estimates seem robust across
simulation conditions. There is no clear effect of the number of
iterations on the bias in these estimates, while the convergence
diagnostics indicate that the algorithm did not reach a completely
converged state (yet).

\hypertarget{multivariate-estimates}{%
\subsection{Multivariate estimates}\label{multivariate-estimates}}

\textbf{This is only true for 75\% missingness:} There is a clear bias
in the regression estimates in simulation conditions where the number of
iterations is smaller than four. In simulation conditions where
\(T > 5\) there is little to no bias in the estimated regression
coefficients. In most simulation conditions nominal coverage is obtained
(i.e., coverage rates of.95) for the confidence intervals around the
regression coefficients. Conditions with only one or two iterations show
some under-coverage. Since confidence interval width is stable across
conditions, the under-coverage may be attributed to the bias in the
estimated regression coefficients.

\textbf{This is only true for 75\% missingness:} If we look at the
estimated proportion of explained variance in outcome variable Y we see
that the coefficient of determination is underestimate estimated in
conditions where the number of iterations is equal to two or less, and
slightly overestimated in conditions where the number of iterations is
equal to three or more.

In short, we see that the minimum number of iterations required to
obtain unbiased, confidence valid regression estimates is 5. This value,
however, is dependent on the percentage of missing values. E.g., with
95\% of cases having missing data we need at least seven iterations to
obtain unbiased results.

@@

\emph{Proper performance under a `missing completely at random'
missingness mechanism is necessary to demonstrate the appropriateness of
\(\widehat{R}\) and \(AC\) as non-convergence diagnostics. However,
results may not be extrapolated to other missingness mechanisms.}

As expected, complete convergence (\(\widehat{R}=1\) and \(AC=0\)) is
never observed. But what about the thresholds for practice?
\(\widehat{R}\) \textless1.2 is not stringent enough, because conditions
in which \(\widehat{R}\) is smaller than 1.2 do not all yield unbiased
estimates of the \(Q\)s. Moreover, there is a dip in the \(\widehat{R}\)
values, after which \(\widehat{R}\) \textgreater{} 1.2 again. If the
algorithms would be terminated at the iteration where \(\widehat{R}\)
\textless{} 1.2 occurs for the first time, the increase in
\(\widehat{R}\) values might be missed. The threshold \(\widehat{R}\)
\textless{} 1.1 is somewhat conservative in comparison with the
performance measures. Overall it seems OK. \(\widehat{R}\) \textless{}
1.01 is too stringent compared to the performance measures. And it is
not obtained for any number of iterations (\textbf{but may be necessary
in more complex complete data models???}).

This note shows that convergence diagnostics \(\widehat{R}\) and \(AC\)
may diagnose convergence of multiple imputation algorithms, but their
performance differs from conventional applications to iterative
algorithmic procedures. (\textbf{nope! it shows that MICE can lead to
correct outcomes when they have not converged according to two common
conv diags. This may be due to the measures (e.g., assumption of
overdisp) or due to the Qs (lm reg coeff, not higher dimensional/more
complex RQs). Add what \%miss has to do with it.})

\(\widehat{R}\) and autocorrelation indicate that algorithmic
convergence may only be reached after twenty or even forty iterations,
while unbiased, confidence valid estimates may be obtained with as
little as four iterations. These results are in agreement with the
simulation hypothesis: \(\widehat{R}\) over-estimates the severity of
non-convergence when applied to MI procedures. \%This may be due to the
quantity of scientific interest chosen. More `complicated' \(Q\)s (e.g.,
higher-order effects or variance components) might show bias, under- or
over-coverage at higher \(T\).

According to this simulation study, the recently proposed threshold of
\(\widehat{R}<1.01\) may be too stringent for MI algorithms.
(\textbf{This is only one of the goals: to give applied researchers a
diagnostic to indicate that they should keep iterating. The other is the
default in mice and other software packages, and yet another is\ldots{}
i forgot}) Under the relatively easy missing data problem of the current
study, the threshold was not reached. The other extreme of the
\(\widehat{R}\)-thresholds, the conventionally acceptable
\(\widehat{R} <1.2\), may be too lenient for MI procedures. Applying
this threshold to the current data, lead to falsely diagnosing
convergence at \(T = 3\) (\textbf{because it goes up after, not because
it is not converged enough}). It appears that the widely used threshold
of \(\widehat{R} < 1.1\) suits MI algorithms the best. We might,
however, also formulate a new threshold, specifically for the evaluation
of MI algorithms. The current study suggests that \(\widehat{R} < 1.05\)
may be implemented, since that is the level at which the \(\widehat{R}\)
stabilize (around \(T = 20\)) (\textbf{Not necessary for this Q, but
maybe for more complicated Qs}).

The negative \(AC\)-values obtained in this study show no threat of
non-stationarity. However, the initial dip in \(AC\)-values
\textbf{(which disappeared!)} may have implications for the default
number of iterations in \texttt{mice} (\texttt{maxit\ =\ 5}).
Terminating the algorithm at \(T=5\) may not be the most appropriate,
since this lead to the worst convergence (\textbf{nope, only for Rh, not
\(AC\)}), as indicated by \(\widehat{R}\) and \(AC\). Under the current
specifications, \(T>20\) would be more appropriate.

Further research is needed to investigate their performance under clear
violation of convergence, e.g.~dependency between predictors (predictors
with very high correlations). Until then, we have only shown that the
convergence diagnostics can diagnose non-convergence of MI algorithms
that trend towards a converged state. Also for future research, look at
developing a convergence diagnostic for substantive models, and
implement a Wald test for \(AC = 0\).

@@

For empirical researchers: 1) Check trace plots for pathological non
convergence and adjust imputation model if necessary. 2) use
\(\widehat{R}\) wait 1.1 ash threshold and autocorrelation with
\textbf{?} as threshold. Keep iterating until these thresholds are
reached. 3) Do not use the R function ACF. Instead, compute
autocorrelations manually \textbf{(see e.g., {[}GitHub link{]})}. 4)
Track your own scalar summary of interest. This is somewhat advanced but
explained in Van Buren 2018. Compute \(\widehat{R}\) and autocorrelation
values for this scalar summary. 5) \textbf{Something} about the novel
\(\theta\) that is `substantive model-independent'.

@@

\textbf{Pmm, M(N)AR, Empirical data, Vignette, ShinyMICE, AC across
imputations, not iterations (new convergence diagnostic unique for
iterative imputation??).}

\textbf{If I don't include it in my study, also: Significance of \(AC\)
values.}

@@

Scientific contribution (clear research question, convincing
argumentation that the study adds to the scientific literature)

Theory (use of theory, clear hypotheses, relevant literature)

Method (correct data collection/handling, measures, choice and
description of method of analysis)

Results (correct interpretation of findings, clear presentation)
Discussion and conclusions (summary, broader implications, limitations,
future research)

Written presentation (writing, structure, tables and figures,
references)

>>>>

We want to know what the effect is of inducing non-conv of the bias in
all Qs and the confidence validity of the betas. So we plot the bias for
each Q against the number of iterations and the missingness proportion
in the condition. We look at averages across simulation repetitions. Is
there more bias in conditions where non-conv was induced? (NOPE for
univariates, YES for multivar.).

Then, we want to know if conditions in which the Qs are affected by
non-convergence (on average) are also identified by the conv diag. We do
this by looking at the bias in Q against the conv diag applied to
related thetas. Is the point at which there is no more bias the same
point at which the conv diag indicate minimal signs of non-conv? (NOPE,
they are too conservative).

Is one of the thetas better? (NAH, doesn't really matter).

What should the thresholds be? (LOWER than recommended) And the default
nr of iterarions? (MAYBE increase to 10?)

\textbf{Add to this section: 1) emphasize that the plots are averages
across repetitions, not within MICE; 2) sample effects due to single
complete dataset. 3) Add more info about figure legends and axes.)}

For reasons of brevity, we only discuss the convergence diagnostics for
the \(Q\) with the worst performance in terms of bias. For the
descriptve statistics, the magnitude of the bias was the largest in
\(Y\) (i.e, \(j = Y\) in \(\mu_j\) and \(\sigma_j\)). For the
\(Q=\beta_{1,2,3}\), \(\beta_1\) showed the worst perfomance (i.e, the
effect of \(X_1\) on \(Y\). Moreover, we only show conditions where
\(T\leq50\) in the figures. Full results are available from
\textbf{{[}link github repo{]}}.

\hypertarget{in-general}{%
\subsection{In general}\label{in-general}}

@@

Evaluation with the performance measures shows that \(\widehat{R}\) is
conservative: it identifies signs of non-convergence in conditions where
the \(\bar{Q}\)s are unbiased and/or confidence valid estimates of the
\(Q\)s.

\hypertarget{means}{%
\subsection{Means}\label{means}}

\textbf{Message: after one iteration (T=2), the algorithm is as well
converged as the missingness proportion allows. Sooner even with lower
missingness proportions (T=1 might be ok for p\_mis=5 and 25).} Number
of it has little influence on bias in hat\{Q\}, only in first iteration
some bias in missingness conditions of 25 and up. There is a clear
different between the missingness conditions. The higher the missingness
prop, the larger the magnitude of the bias. For this Q, t=2 would be
sufficient. And only for p\_mis = 5 the bar\{Q\}s are unbiased. All
other p\_mis conditions have bias irrespective of the number of
iterations.

Rhat differs a lot across iteration conditions. But not structurally for
missingness proportions. Highest Rhats for the condition t=2, while the
most bias is for (t=1), but there is no Rhat for that by definition. But
for missingness, 25\% seems the most signs of non-conv, while it has one
of the least bias. Threshold of 1.2 reached in condition it = 3, p\_mis
= 5 and 75, and then for it\textgreater6, p\_mis= 5 and 50, and
it\textgreater10 for all p\_mis. Threshold 1.1 reached in in it=12,
p\_mis = all but 75, 1.01 only reached in it = 93, p\_mis = 75. While
for this Q, t=2 would be sufficient.

AC mainly shows effects of the set of iteration conditions. AC indicates
less signs of non-conv in conditions with higher number of iterations.
For iteration conditions where t\textless10, there are also some effects
of the missingness proportions. The p\_mis condtions 25 and 95 \%
indicate the most signs of non-convergence. For p\_mis = 95\% this
corresponds to the bias in Qbar. For 25\% miss it doesn't. The
thresholds defined by the critical value is only met in the conditions
it=3, p\_mis = 5,75, and 95\%.

@@

\hypertarget{standard-deviations}{%
\subsection{Standard deviations}\label{standard-deviations}}

\textbf{Message: even with possible effect of initial values (T=1), the
algorithm is as well converged as the missingness proportion allows.}
For \(Q=\sigma_Y\) there is no effect visible of the iteration
conditions, but clearly for the p\_mis conditions. Most bias for p\_mis
= 95, followed by p\_mis=50 and 75. For the least possible bias we may
use any number of iterations \(T\geq1\), but none of the p\_mis results
in unbiased bar\{Q\}s.

There is a clear effect of the number of iterations on Rhat. Conditions
with lower Ts have more signs of non-conv, as identified by Rhat. There
is little difference between missingness conditions. Only for conditions
where \(T<5\) the missingness proportions seem to affect whether Rhat
indicates non-conv. The bias in bar\{Q\} seems unrelated to the
magnitude of Rhat. Conditions with more bias (p\_mis = 50,75,95) are not
identified to have more signs of non-convergence. The thresholds are
therefore useless (because depend mostly on t, not so much p\_mis).

AC is affected by both t and p\_mis conditions. But conditions with
worst bias (p\_mis = 50, 75, 95) are not identified as worst converging
according to AC. Moreover, only for T=3 the AC is significantly
different from zero.

@@

\hypertarget{regression-coefficients}{%
\subsection{Regression coefficients}\label{regression-coefficients}}

\textbf{Message: after four iterations (T=5), the algorithm is as well
converged as the missingness proportion allows. Sooner even with lower
missingness proportions (T=1 might be ok for p\_mis=5 and 25).} The
figure for \(Q=\beta_1\) shows that the \(\bar{Q}\)s are affected by
both the number of iterations and the missingness proportion. For T=1,
the magnitude of the bias (almost) follows the order of the missingness
proportions. More missingness causes more bias in the \(\bar{Q}\). In
general, a lower nr of iterations results in more bias, but this depends
on the missingness proportion. For p\_mis=95, there is more bias in
conditions where T\textless6. When p\_mis = 50, it's T\textless3. For
p\_mis = 75 and 25, it's only for T\textless2. And conditions where
p\_mis = 5 are unaffected on average. However, none of the conditions
result in completely unbiased estimates of \(Q\).

We should see this in the conv diag: higher Rhat and AC for p\_mis=75
and 95, and for condtions where \(T<2-6\), depending on p\_mis.

Rhat depends somewhat on the missingness proportion. For conditions
where T\textless10, the Rhat values follow the p\_mis more or less. But
what we actually want to know is whether they are indicators of bias.
Conditions with the worst bias (high p\_mis, low T) have the largest
Rhat values. The thresholds are not appropriate to diagnose sufficient
convergence. All thresholds over-estimate the signs of non-convergence
compared to the bias in Qbar. However, a higher threshold than 1.2 may
result in terminating the algorithm after 3 iterations, which is
insufficient for the p\_mis = 95 condition. Threshold of 1.25 or 1.3
could work for the other p\_mis conditions.

AC is very affected by the p\_mis. The two conditions with the most bias
(p\_mis = 75 and 95) are the only ones identified as non-converging, as
they should. However, the p\_mis = 75 condition is only identified in
the T=3 condtion, while the bias is the least pronounced there. And the
AC only identifies non-convergence in the p\_mis = 95 conditions where
T\textgreater10. In the default situation, T=5, this would be
overlooked. Also, the AC shows the opposite of what you would expect
from a theta that is based on the observed data: shouldn't there be
higher autocorrelations in conditions with comparatively more observed
data in the completed datasets? I don't understand this.

\textbf{Add how interesting it is because we literally track Q and are
still not ok.}

@@

\hypertarget{variance-explained}{%
\subsection{Variance explained}\label{variance-explained}}

\textbf{Message: after two iterations (T=3), the algorithm is as well
converged as the missingness proportion allows. Sooner even with lower
missingness proportions (T=1 might be ok for p\_mis=5 and 25).} Bias is
the worst for conditions where p\_mis = 95, T\textless3. Followed by
p\_mis = 75, T=1; and p\_mis = 50, T=1. Conditions where p\_mis = 25 or
5 seem unaffected by the number of iterations. Minimal bias is obtained
in all iteration conditions. We should see this in Rhat and AC: worst
convergence in p\_mis = 95 and 50; some non-conv in conditions where
p\_mis = 75 and T=1, and no non-conv in conditions where p\_mis = 25 or
5. The sign of the bias is logical: when there is less info in the data,
the relations between the predictors and the outcome is under-estimated.

Rhat behaves oppositely to what we wanted. The lowest Rhat values for
conditions where T\textless10 are observed for the two missingness
conditions with the worst bias. However, the highest Rhat observed is
indeed for the missingness condition with the worst bias, p\_mis =95.
But only at T=5 (while bias is no worse than anywhere else in this
p\_mis condition, conditional on iterations condition T\textgreater2).
So neither of the thresholds works perfectly.

AC shows a similar trend to the biases: more signs of non-conv for
higher p\_mis and lower T. The threshold, however, performs less well.
It identifies signs of non-convergence for all conditions where \(T=3\),
irrespective of p\_mis. Then nothing, and then for p\_mis =95,
T\textgreater7, and p\_mis=50, T \textgreater14, and p\_mis = 75,
T\textgreater20. This does not corresond to the bias in \(\bar{Q}\).
Maybe use AC\textgreater.5 at T=5 as threshold?

@@

@@

In short, univariate estimates seem robust against terminating the
algorithm early: There is no clear effect of the number of iterations on
the bias in these estimates. The bias in the Qs only depends on the
p\_mis. Yet, the convergence diagnostics indicate that the algorithm did
not reach a completely converged state (yet).

\(\widehat{R}\) and autocorrelation indicate that algorithmic
convergence may only be reached after twenty or even forty iterations,
while unbiased, confidence valid estimates may be obtained with as
little as one iteration. These results are in agreement with the
simulation hypothesis: \(\widehat{R}\) over-estimates the severity of
non-convergence when applied to MI procedures.

With that, we show that MICE can lead to correct outcomes when they have
not converged according to two common conv diags. This may be due to the
methods (and their thresholds) or due to the Qs (descriptives and
multivariate linear regression, not higher dimensional/more complex Qs).
More `complicated' \(Q\)s (e.g., higher-order effects or variance
components) might show bias, under- or over-coverage at higher \(T\), as
indiated by Rhat and AC. Add what \%miss has to do with it.

@@

For empirical researchers:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Check trace plots for pathological non convergence and adjust
  imputation model if necessary.
\item
  use \(\widehat{R}\) wait 1.1 ash threshold and autocorrelation with
  \textbf{?} as threshold. Keep iterating until these thresholds are
  reached.
\item
  Do not use the R function ACF. Instead, compute autocorrelations
  manually \textbf{(see e.g., {[}GitHub link{]})}.
\item
  Track your own scalar summary of interest. This is somewhat advanced
  but explained in Van Buren 2018. Compute \(\widehat{R}\) and
  autocorrelation values for this scalar summary.
\item
  \textbf{Something} about the novel \(\theta\) that is `substantive
  model-independent'.
\end{enumerate}

@@

Further research is needed to investigate their performance under clear
violation of convergence, e.g.~dependency between predictors (predictors
with very high correlations). Also:

\begin{itemize}
\item
  Induce non-convergence with mis-specified imputation model.
\item
  More difficult missingness problems, i.e., M(N)AR instead of MCAR.
  Proper performance under a `missing completely at random' missingness
  mechanism is necessary to demonstrate the appropriateness of
  \(\widehat{R}\) and \(AC\) as non-convergence diagnostics. However,
  results may not be extrapolated to other missingness mechanisms.
\item
  Imputation technique with questionable convergence properties, e.g.,
  PMM.
\item
  Apply recommendations to empirical data and write a vignette for
  applied researchers.
\item
  Implement recommendations in ShinyMICE.
\item
  Develop an additional convergence diagnostic unique for iterative
  imputation: AC across imputations, instead of iterations.
\end{itemize}

@@

\begin{itemize}
\item
  Scientific contribution (clear research question, convincing
  argumentation that the study adds to the scientific literature)
\item
  Theory (use of theory, clear hypotheses, relevant literature)
\item
  Method (correct data collection/handling, measures, choice and
  description of method of analysis)
\item
  Results (correct interpretation of findings, clear presentation)
\item
  Discussion and conclusions (summary, broader implications,
  limitations, future research)
\item
  Written presentation (writing, structure, tables and figures,
  references)
\end{itemize}
